Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond

### 2024-04-11 main.tex
:dizzy: Geodesic RCG

@@

\begin{document}
\title{Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{yang.zhao18, c.li21, b.clerckx\}@imperial.ac.uk).

		Massimo Franceschetti is with the Department of Electrical and Computer Engineering, University of California at San Diego, La Jolla CA 92093, USA (e-mail: massimo@ece.ucsd.edu).
	}
}
\maketitle

\begin{abstract}
	This paper investigates how a passive \gls{ris} can manipulate the \gls{mimo} point-to-point channel in terms of singular values and power gain.
	We depart from the widely-adopted diagonal phase shift model to a \gls{bd} architecture, which enables {cooperative} wave scattering thanks to in-group connections between elements.
	An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise {geodesic} \gls{rcg} in the Stiefel manifold, which performs multiplicative rotational (instead of add-then-retract) updates for faster convergence.
	% For the singular value shaping problem, we numerically characterize their Pareto frontiers by weighted sum optimization via geodesic \gls{rcg}, which generalizes most relevant metrics (e.g., degree of freedom and condition number).
	% Many analytical singular value bounds are also derived for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	For the singular value problem, we numerically characterize their Pareto frontiers by geodesic \gls{rcg} and derive analytical bounds for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	% For the singular value problem, we consider a numerical optimization approach that characterizes the Pareto frontiers as well as an analytical bounding approach in special (i.e., rank-deficient and fully-connected) scenarios.
	For the power gain problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% Those channel shaping designs have a two-fold advantage.
	Those channel shaping designs have two major implications.
	First, they demystify the performance gain from off-diagonal entries in terms of {subchannel rearrangement} and {subspace alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in subchannel rearrangement and subspace alignment.
	Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} approach and a two-stage shaping-inspired approach.
	Simulation results indicate that the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal towards unitary, while the relative gain of \gls{bd}-\gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

@@

	While the above works offer promising glimpses into the channel shaping potential of passive \gls{ris}, they neither suggest the achievable singular value region nor address the most general \gls{bd}-\gls{ris}-aided \gls{mimo} scenario with direct link and arbitrary group size.
	On the other hand, the great potential of \gls{bd}-\gls{ris} (especially in \gls{mimo}) has not been fully understood and an efficient design framework is still missing.
	This paper aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

@@

	Second, we exploit the Riemannian geometry of the Stiefel manifold and propose an efficient \gls{bd}-\gls{ris} design framework based on geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	This method modified from \cite{Abrudan2008,Abrudan2009} not only features lower complexity and faster convergence than general non-geodesic approach \cite{Absil2009,Pan2022d}, but also works for arbitrary group size and any smooth optimization problem.
	Specifically, block-wise multiplicative rotational updates are performed along the geodesics of the Stiefel manifold and compactly evaluated as the exponential map \cite{Edelman1998}.
	By exploiting the inherent structure of unitary matrices, this method avoids retractions from the Euclidean space and improves the computational efficiency and stability.
	This is the first work to tailor an efficient and universal optimization framework for \gls{bd}-\gls{ris}.

@@

	% Second, we propose a \gls{rcg} algorithm adapted from \cite{Abrudan2008,Abrudan2009} for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size.
	% Specifically, block-wise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold, which are expressed compactly by the exponential map \cite{Edelman1998}.
	% It features lower complexity and faster convergence than general manifold optimization \cite{Absil2009,Pan2022d}, and solves the Pareto singular value problem.
	% This is the first paper to tailor an efficient optimization framework for asymmetric \gls{bd}-\gls{ris}.

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that further divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis and explore the performance limits, we assume a lossless asymmetric network without mutual coupling between scattering elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is block-unitary
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is the $g$-th unitary block (i.e., $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$) that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	Note that diagonal and unitary \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the radiation pattern and circuit topology should be modelled in the scattering matrix.

@@

	\begin{remark}
		By analyzing \eqref{eq:channel_equivalent_svd}, we conclude that the off-diagonal entries of the \gls{bd}-\gls{ris} scattering matrix provide two key potentials for \gls{mimo} channel shaping:
		\begin{itemize}
			\item \textbf{Subchannel rearrangement:} This unique feature of \gls{bd}-\gls{ris} allows each group to rearrange and recombine the backward and forward subchannels by strength. In \gls{siso}, diagonal \gls{ris} with perfect phase matching provides a maximum indirect channel amplitude of $\sum_{n=1}^{N_\mathrm{S}} \lvert h_{\mathrm{B},n} \rvert \lvert h_{\mathrm{F},n} \rvert$, while \gls{bd}-\gls{ris} can generalize it to $\sum_{g=1}^{G} \sum_{l=1}^{L} \lvert h_{\mathrm{B},\pi_{\mathrm{B},g}(l)} \rvert \lvert h_{\mathrm{F},\pi_{\mathrm{F},g}(l)} \rvert$, where $\pi_{\mathrm{B},g}$ and $\pi_{\mathrm{F},g}$ are permutations of $\mathcal{L} \triangleq \{1, \ldots, L\}$.
			Note the first summation is over groups and the second summation is over permuted subchannels.
			By rearrangement inequality, the maximum is attained by pairing the $l$-th strongest backward and forward subchannels within each group.
			Since the number of subchannels associated with each group is proportional to $N_\mathrm{T} N_\mathrm{R}$, we conclude the advantage of \gls{bd}-\gls{ris} in subchannel rearrangement scales with \gls{mimo} dimensions,
			\item \textbf{Subspace alignment:} Each group can align the singular vectors of the associated backward-forward (intra-group, multiplicative) channels and direct-indirect (inter-group, additive) channels. In \gls{siso}, subspace alignment boils down to phase matching and the optimal scattering matrix of group $g$ that maximizes the channel gain is
			\begin{equation}
				\mathbf{\Theta}_g^\star = \exp \bigl(\jmath \mathrm{arg}(h_\mathrm{D})\bigr) \mathbf{V}_{\mathrm{B},g} \mathbf{U}_{\mathrm{F},g}^\mathsf{H},
			\end{equation}
			where $\mathbf{V}_{\mathrm{B},g} = \bigl[\mathbf{h}_{\mathrm{B},g}/\lVert \mathbf{h}_{\mathrm{B},g} \rVert, \mathbf{N}_{\mathrm{B},g}\bigr] \in \mathbb{U}^{L \times L}$, $\mathbf{U}_{\mathrm{F},g} = \bigl[\mathbf{h}_{\mathrm{F},g}/\lVert \mathbf{h}_{\mathrm{F},g} \rVert, \mathbf{N}_{\mathrm{F},g}\bigr] \in \mathbb{U}^{L \times L}$, and $\mathbf{N}_{\mathrm{B},g}, \mathbf{N}_{\mathrm{F},g} \in \mathbb{C}^{L \times (L-1)}$ are the orthonormal bases of the null spaces of $\mathbf{h}_{\mathrm{B},g}$ and $\mathbf{h}_{\mathrm{F},g}$, respectively.
			Diagonal \gls{ris} ($L=1$, empty null spaces) thus suffices for perfect phase matching in \gls{siso}.
			When it comes to \gls{mimo}, each individual scattering element can only apply a common phase shift to the ``pinhole'' indirect channel $\mathbf{H}_g \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{T}}$ passing through itself.
			That is, the disadvantage of diagonal \gls{ris} in subspace alignment scales with \gls{mimo} dimensions.
			As will be shown later, even if the \gls{bd}-\gls{ris} is fully-connected, there still exists a tradeoff between the alignment of direct-indirect and backward-forward subspaces.
		\end{itemize}
	\end{remark}

@@

\begin{section}{Channel Singular Values Redistribution}
	\begin{subsection}{A Toy Example}\label{sc:toy_example}
		We first illustrate the channel shaping capabilities of different \gls{ris} by a toy example.
		% We first use a toy example to illustrate that \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values.
		Consider a $2 \times 2 \times 2$ setup where the direct link is blocked.
		The diagonal \gls{ris} is modeled by $\mathbf{\Theta}_\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the unitary \gls{bd}-\gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}_\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
		\end{equation}
		In particular, $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		We also enforce symmetry by $\beta = \pi / 2$ such that both architectures have the same number of angular parameters.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/singular_trend.eps}
			\caption{$2 \times 2 \times 2$ (no direct) channel singular value shaping by diagonal and symmetry unitary \gls{ris}.}
			\label{fg:singular_trend}
		\end{figure}
		Fig.~\ref{fg:singular_trend} shows the channel singular values achieved by an exhaustive grid search over $(\theta_1, \theta_2)$ for diagonal \gls{ris} and $(\alpha, \psi)$ for symmetric unitary \gls{ris}.
		It is observed that both singular values can be manipulated up to $9\%$ using diagonal \gls{ris} and $42\%$ using symmetric \gls{bd}-\gls{ris}, despite both architectures have the same number of scattering elements and design parameters.
		A larger performance gap is expected when asymmetric \gls{bd}-\gls{ris} is available.
		This example shows \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values and motivates further studies on channel shaping.
	\end{subsection}

	\begin{subsection}{Pareto Frontier Characterization}
		We then characterize the Pareto frontier of channel singular values by maximizing their weighted sum
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\sum_n \rho_n \sigma_n(\mathbf{H})}{\label{op:pareto}}{\label{ob:pareto}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I},}{\quad \forall g,}{\label{co:pareto_unitary}}
		\end{maxi!}
		where $n \in \{1,\ldots,N \triangleq \min(N_\mathrm{T}, N_\mathrm{R})\}$ and $\rho_n$ is the weight of the $n$-th singular value that can be positive, zero, or negative.
		Varying $\{\rho_n\}$ unveils the entire achievable singular value region.
		Thus, the Pareto frontier problem~\eqref{op:pareto} generalizes most relevant metrics and provides a powerful shaping framework.
		The objective \eqref{ob:pareto} is smooth in $\mathbf{\Theta}$ and the feasible domain \eqref{co:pareto_unitary} for group $g$ corresponds to the Stiefel manifold.
		Next, we zoom out to general smooth maximization problems of asymmetric \gls{bd}-\gls{ris}.

>>>>

% Double superscript
\newcommand\herm[2][-4]{{#2}^{\mkern#1mu\mathsf{H}}}

@@

\begin{document}
\title{Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

		Massimo Franceschetti is with the Department of Electrical and Computer Engineering, University of California at San Diego, La Jolla CA 92093, USA (e-mail: \href{mailto:massimo@ece.ucsd.edu}{massimo@ece.ucsd.edu}).
	}
}
\maketitle

\begin{abstract}
	This paper investigates how a passive \gls{ris} can manipulate the \gls{mimo} point-to-point channel in terms of singular values and power gain.
	We depart from the widely-adopted diagonal phase shift model to a \gls{bd} architecture, which enables {cooperative} wave scattering thanks to in-group connections between elements.
	An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise {geodesic} \gls{rcg} on the Stiefel manifold, which performs multiplicative rotational (instead of add-then-retract) updates for faster convergence.
	% For the singular value shaping problem, we numerically characterize their Pareto frontiers by weighted sum optimization via geodesic \gls{rcg}, which generalizes most relevant metrics (e.g., degree of freedom and condition number).
	% Many analytical singular value bounds are also derived for rank-deficient channels and unitary (a.k.a. unitary) \gls{ris}.
	For the singular value problem, we numerically characterize their Pareto frontiers by geodesic \gls{rcg} and derive analytical bounds for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	% For the singular value problem, we consider a numerical optimization approach that characterizes the Pareto frontiers as well as an analytical bounding approach in special (i.e., rank-deficient and fully-connected) scenarios.
	For the power gain problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% Those channel shaping designs have a two-fold advantage.
	Those channel shaping designs have two major implications.
	First, they demystify the performance gain from off-diagonal entries in terms of {subchannel rearrangement} and {subspace alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in subchannel rearrangement and subspace alignment.
	Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} approach and a two-stage shaping-inspired approach.
	Simulation results indicate that the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal towards unitary, while the relative gain of \gls{bd}-\gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

@@

	While the above works offer promising glimpses into the channel shaping potential of passive \gls{ris}, they neither suggest the achievable singular value region nor address the most general scenario with direct link and arbitrary group size.
	On the other hand, \gls{bd}-\gls{ris} has been mainly investigated for beamforming and its channel shaping potential in \gls{mimo} has not been fully understood.
	% the great potential of \gls{bd}-\gls{ris} (especially in \gls{mimo}) has not been fully understood and an efficient design framework is still missing.
	This paper aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

@@

	Second, we exploit the Riemannian geometry of the Stiefel manifold and propose an efficient \gls{bd}-\gls{ris} design framework based on geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	This method modified from \cite{Abrudan2008,Abrudan2009} not only provides better objective value and faster convergence than general non-geodesic approach \cite{Absil2009,Pan2022d}, but also works for arbitrary group size and any smooth optimization problem.
	Specifically, group-wise multiplicative rotational updates are performed along the geodesics of the Stiefel manifold and compactly evaluated as the exponential map \cite{Edelman1998}.
	By exploiting the inherent structure of unitary matrices, this method avoids retractions from the Euclidean space and improves the computational efficiency and stability.
	This is the first work to tailor an efficient and universal optimization framework for \gls{bd}-\gls{ris}.

@@

	% Second, we propose a \gls{rcg} algorithm adapted from \cite{Abrudan2008,Abrudan2009} for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size.
	% Specifically, group-wise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold, which are expressed compactly by the exponential map \cite{Edelman1998}.
	% It features lower complexity and faster convergence than general manifold optimization \cite{Absil2009,Pan2022d}, and solves the Pareto singular value problem.
	% This is the first paper to tailor an efficient optimization framework for asymmetric \gls{bd}-\gls{ris}.

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that further divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis and explore the performance limits, we assume a lossless asymmetric network without mutual coupling between scattering elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is block-unitary
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
		\label{eq:bd_ris}
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is the $g$-th unitary block (i.e., $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$) that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	Note that diagonal and unitary \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the radiation pattern and circuit topology need to be modelled in the scattering matrix.

@@

	% \begin{remark}
	By analyzing \eqref{eq:channel_equivalent_svd}, we conclude that the off-diagonal entries of the \gls{bd}-\gls{ris} scattering matrix provide two key potentials for \gls{mimo} channel shaping:
	\begin{itemize}
		\item \emph{Subchannel rearrangement:} This unique feature of \gls{bd}-\gls{ris} allows each group to rearrange and recombine the backward and forward subchannels by their strength. In \gls{siso}, diagonal \gls{ris} with perfect phase matching provides a maximum indirect channel amplitude of $\sum_{n=1}^{N_\mathrm{S}} \lvert h_{\mathrm{B},n} \rvert \lvert h_{\mathrm{F},n} \rvert$, while \gls{bd}-\gls{ris} can generalize it to $\sum_{g=1}^{G} \sum_{l=1}^{L} \lvert h_{\mathrm{B},\pi_{\mathrm{B},g}(l)} \rvert \lvert h_{\mathrm{F},\pi_{\mathrm{F},g}(l)} \rvert$, where $\pi_{\mathrm{B},g}$ and $\pi_{\mathrm{F},g}$ are permutations of $\mathcal{L} \triangleq \{1, \ldots, L\}$.
		Note the first summation is over groups and the second summation is over permuted subchannels.
		By rearrangement inequality, the maximum is attained by pairing the $l$-th strongest backward and forward subchannels within each group.
		Since the number of subchannels associated with each group is proportional to $N_\mathrm{T} N_\mathrm{R}$, we conclude the advantage of \gls{bd}-\gls{ris} in subchannel rearrangement scales with \gls{mimo} dimensions,
		\item \emph{Subspace alignment:} Each group can align the singular vectors of the associated backward-forward (intra-group, multiplicative) channels and direct-indirect (inter-group, additive) channels. In \gls{siso}, subspace alignment boils down to phase matching and the optimal scattering matrix of group $g$ that maximizes the channel gain is
		\begin{equation}
			\mathbf{\Theta}_g^\star = \exp \bigl(\jmath \mathrm{arg}(h_\mathrm{D})\bigr) \mathbf{V}_{\mathrm{B},g} \mathbf{U}_{\mathrm{F},g}^\mathsf{H},
		\end{equation}
		where $\mathbf{V}_{\mathrm{B},g} = \bigl[\mathbf{h}_{\mathrm{B},g}/\lVert \mathbf{h}_{\mathrm{B},g} \rVert, \mathbf{N}_{\mathrm{B},g}\bigr] \in \mathbb{U}^{L \times L}$, $\mathbf{U}_{\mathrm{F},g} = \bigl[\mathbf{h}_{\mathrm{F},g}/\lVert \mathbf{h}_{\mathrm{F},g} \rVert, \mathbf{N}_{\mathrm{F},g}\bigr] \in \mathbb{U}^{L \times L}$, and $\mathbf{N}_{\mathrm{B},g}, \mathbf{N}_{\mathrm{F},g} \in \mathbb{C}^{L \times (L-1)}$ are the orthonormal bases of the null spaces of $\mathbf{h}_{\mathrm{B},g}$ and $\mathbf{h}_{\mathrm{F},g}$, respectively.
		Diagonal \gls{ris} ($L=1$, empty null spaces) thus suffices for perfect phase matching in \gls{siso}.
		When it comes to \gls{mimo}, each individual scattering element can only apply a common phase shift to the ``pinhole'' indirect channel $\mathbf{H}_g \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{T}}$ passing through itself.
		That is, the disadvantage of diagonal \gls{ris} in subspace alignment scales with \gls{mimo} dimensions.
		As will be shown later, even if the \gls{bd}-\gls{ris} is unitary, there still exists a tradeoff between the alignment of direct-indirect and backward-forward subspaces.
	\end{itemize}
	% \end{remark}

@@

\begin{section}{Group-Wise Geodesic \glsfmtshort{rcg}}
	In this section, we first provide an overview on signal processing techniques for general \gls{bd}-\gls{ris} design problems, then propose a novel group-wise geodesic \gls{rcg} method that exploits the properties of unitary group to operate directly on the Stiefel manifold.
	% The proposed method is applicable to any smooth optimization problem on the Stiefel manifold and is particularly suitable for \gls{bd}-\gls{ris} design problems.
	We will later show that the proposed method not only provides better objective value and faster convergence than other approaches, but also works for arbitrary group size and any smooth optimization problem.
	\begin{subsection}{Conventional Techniques}
		General \gls{bd}-\gls{ris} design techniques can be classified into two categories based on the optimization variable and hardware architecture:
		\begin{itemize}
			\item \emph{Scattering matrix $\mathbf{\Theta}$:} This approach is often exploited for asymmetric architectures where the feasible domain of each group is a $L$-dimensional Stiefel manifold $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$. Due to its non-convexity, relevant problems are usually solved by general non-geodesic manifold optimization \cite{Li2023b,Li2023c,Bartoli2023} or relax-then-project methods \cite{Fang2023}. The former will be discussed in the next subsection.
			\item \emph{Reactance matrix $\mathbf{X}$:} This approach is often exploited for symmetric architectures where every pair of elements in the same group are connected by capacitors and inductors. According to network theory \cite{Pozar2011}, it maps to the scattering matrix by $\mathbf{\Theta}_g = (\jmath \mathbf{X}_g + Z_0 \mathbf{I})^{-1} (\jmath \mathbf{X}_g - Z_0 \mathbf{I})$, which formulates an unconstrained optimization problem on the upper triangular entries of $\{\mathbf{X}_g\}_{g \in \mathcal{G}}$ that is solvable by quasi-Newton methods \cite{Shen2020a}.
		\end{itemize}
	\end{subsection}

	\begin{subsection}{Geodesic vs Non-Geodesic \glsfmtshort{rcg}}
		We first revisit the general non-geodesic \gls{rcg} method that is applicable to optimization problems over arbitrary manifolds \cite{Absil2009,Pan2022d}.
		The main idea is to perform group-wise \emph{additive} updates along the conjugate direction [not in, but guided by] the tangent space, then project the solution back onto the manifold.
		% update the feasible point by progressing along conjugate directions then project back to the manifold.
		% The main idea is to update the feasible point by progressing along conjugate directions then project back to the manifold.
		For maximization problem with smooth objective $f$ and block-unitary constraint \eqref{eq:bd_ris}, the steps for group $g$ at iteration $r$ are summarized below:
		\begin{enumerate}
			\item \emph{Compute the Euclidean gradient \cite{Hjorungnes2007}:} The gradient of $f$ with respect to $\mathbf{\Theta}_g^*$ in the Euclidean space is
			\begin{equation}
				\nabla_{\mathrm{E},g}^{(r)} = \frac{\partial f(\mathbf{\Theta}_g^{(r)})}{\partial \mathbf{\Theta}_g^*};
				\label{eq:gradient_euclidean}
			\end{equation}
			\item \emph{Translate to the Riemannian gradient \cite{Absil2009,Abrudan2008}:} At point $\mathbf{\Theta}^{(r)}$, the Riemannian gradient lies in the tangent space of the Stiefel manifold $\mathcal{T}_{\mathbf{\Theta}_g^{(r)}}\mathbb{U}^{L \times L} \triangleq \{\mathbf{M} \mid \mathbf{M}^\mathsf{H} \mathbf{\Theta}_g^{(r)} + {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} \mathbf{M} = \mathbf{0}\}$. It can be obtained by projecting the Euclidean gradient onto the tangent space:
			\begin{equation}
				% \nabla_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{E},g}^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)}}^\mathsf{H};
				\nabla_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{E},g}^{(r)} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)}}^\mathsf{H} \mathbf{\Theta}_g^{(r)};
				\label{eq:gradient_riemannian}
			\end{equation}
			\item \emph{Determine the conjugate direction \cite{Nocedal2006,Abrudan2009}:} The conjugate direction is updated over the Riemannian gradient and previous direction as
			\begin{equation}
				\mathbf{D}_g^{(r)} = \nabla_{\mathrm{R},g}^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)}, % $\mathcal{O}(N_\mathrm{S}^2)$
				\label{eq:direction_cg}
			\end{equation}
			where $\gamma_g^{(r)}$ is the deviation parameter that accelerates the convergence. One common choice is the Polak-Ribierre formula \cite{Hager2006}
			\begin{equation}
				\gamma_g^{(r)} = \frac{\mathrm{tr}\bigl((\nabla_{\mathrm{R},g}^{(r)} - \nabla_{\mathrm{R},g}^{(r-1)}) {\nabla_{\mathrm{R},g}^{(r)}}^\mathsf{H}\bigr)}{\mathrm{tr}\bigl(\nabla_{\mathrm{R},g}^{(r-1)} {\nabla_{\mathrm{R},g}^{(r-1)}}^\mathsf{H}\bigr)}; % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
				\label{eq:parameter_cg}
			\end{equation}
			\item \emph{Update [aided by] the tangent space}
		\end{enumerate}
		% It is worth noting that the objective \eqref{ob:pareto} is smooth in $\mathbf{\Theta}$ and the feasible domain \eqref{co:pareto_unitary} corresponds to the Stiefel manifold for group $g$.

		% Both properties are often satisfied in \gls{bd}-\gls{ris} optimization problems, and there exists two general solutions in the literature:
		% \begin{itemize}
		% 	\item \emph{Quasi-Newton method \cite{Shen2020a}:}
		% 	% This is a general-purpose optimization algorithm that approximates the Hessian matrix by the gradient and iteratively updates the scattering matrix along the negative gradient direction.
		% \end{itemize}

		% have exploited either the quasi-Newton method on the reactance
		% Instead of solving it directly, we zoom out to general \gls{bd}-\gls{ris} optimization problem and propose an efficient and universal solution.
		% Next, we zoom out to general smooth maximization problems of asymmetric \gls{bd}-\gls{ris}.

		Inspired by \cite{Abrudan2008,Abrudan2009}, we propose a group-wise \gls{rcg} algorithm along the geodesics on the Lie group of unitary matrices $\mathbb{U}^{L \times L}$.
		It leverages the fact that unitary matrices are closed under multiplication.
		At iteration $r$, the gradient is computed in the Euclidean space and translated to the Riemannian manifold \cite{Absil2009}
		% \begin{gather}
		% 	\nabla_{\mathrm{E},g}^{(r)} = \frac{\partial f(\mathbf{\Theta}_g^{(r)})}{\partial \mathbf{\Theta}_g^*},\label{eq:gradient_euclidean}\\
		% 	\nabla_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{E},g}^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)}}^\mathsf{H}.\label{eq:gradient_riemannian} % $\mathcal{O}(N_\mathrm{S}^3)$
		% \end{gather}
		% The Polak-Ribierre parameter \cite{Polak1969} is approximated as \cite{Abrudan2009}
		% \begin{equation}
		% 	\gamma_g^{(r)} = \frac{\mathrm{tr}\bigl((\nabla_{\mathrm{R},g}^{(r)} - \nabla_{\mathrm{R},g}^{(r-1)}) {\nabla_{\mathrm{R},g}^{(r)}}^\mathsf{H}\bigr)}{\mathrm{tr}\bigl(\nabla_{\mathrm{R},g}^{(r-1)} {\nabla_{\mathrm{R},g}^{(r-1)}}^\mathsf{H}\bigr)}, % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
		% 	\label{eq:parameter_cg}
		% \end{equation}
		% and the conjugate direction is
		% \begin{equation}
		% 	\mathbf{D}_g^{(r)} = \nabla_{\mathrm{R},g}^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)}. % $\mathcal{O}(N_\mathrm{S}^2)$
		% 	\label{eq:direction_cg}
		% \end{equation}
		on the Stiefel manifold, the geodesic emanating from $\mathbf{\Theta}_g^{(r)}$ with velocity $\mathbf{D}_g^{(r)}$ and step size $\mu$ is described compactly by the exponential map \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)}. % $\mathcal{O}(N_\mathrm{S}^3)$
			\label{eq:geodesic}
		\end{equation}
		An appropriate $\mu^\star$ can be obtained by the Armijo rule \cite{Armijo1966}.\footnote{To double the step size, one only need to square the rotation matrix instead of recomputing the matrix exponential, i.e., $\exp(2 \mu \mathbf{D}_g^{(r)}) = \exp^2(\mu \mathbf{D}_g^{(r)})$.}
		Finally, the scattering matrix is updated along the geodesic as
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu^\star).
			\label{eq:update_geodesic}
		\end{equation}
